### 2025-02-03 paper/body_ecml.tex
progress baby

\begin{abstract}
Counterfactual Explanations (CE) have emerged as a popular tool to
explain predictions made by opaque machine learning models: they explain
how factual inputs need to change in order for some fitted model to
produce some desired output. Much existing research has focused on
identifying explanations that are not only valid but also deemed
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning desirable explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and desirable explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver desirable
explanations while maintaining high predictive performance.
\keywords{Counterfactual Explanations \and Explainable
AI \and Representation Learning}
\end{abstract}

@@

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{goodfellow2014explaining} showed that artificial
neural networks (ANN) are sensitive to adversarial examples (AE):
counterfactuals of model inputs that yield vastly different model
predictions despite being semantically indifferent from their factual
counterparts. Despite partially effective mitigation strategies such as
\textbf{adversarial training}, truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards \cite{kolter2023keynote}.

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems involving opaque
models \cite{oneil2016weapons}. Effective XAI tools can aide us in
monitoring models and providing affected individuals with recourse
\cite{wachter2017counterfactual}.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are deemed desirable by humans. Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both desirable and valid for any given
model, independent of whether the learned representations are also
compatible with undesirable explanations \cite{altmeyer2024faithful}.
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity
\cite{schut2021generating}.

@@

\subsection{Background on Counterfactual
Explanations}\label{background-on-counterfactual-explanations}

\cite{wachter2017counterfactual, joshi2019realistic, altmeyer2024faithful}

\subsection{Learning Representations}\label{learning-representations}

\begin{quote}
For example, joint-energy models
\end{quote}

\subsection{Generalization and
Robustness}\label{generalization-and-robustness}

\cite{sauer2021counterfactual} generate counterfactual images for MNIST
and ImageNet through independent mechanisms (IM): each IM learns
class-conditional input distributions over a specific lower-dimensional,
semantically meaningful factor, such as \emph{texture}, \emph{shape} and
\emph{background}. They demonstrate that using these generated
counterfactuals during classifier training improves model robustness.
Similarly, \cite{abbasnejad2020counterfactual} argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on, firstly, identifying similar
input samples with different annotations and, secondly, ensuring that
the gradient of the classifier aligns with the vector between pairs of
counterfactual inputs using the cosine distance as a loss function
(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
be useful for our task as well}}). In the natural language processing
(NLP) domain, counterfactuals have similarly been used to improve models
through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

\subsection{Link to Adversarial
Training}\label{link-to-adversarial-training}

From this perspective, adversarial training induces models to
``unlearn'' representations that are susceptible to the semantically
most meaningless explanations---adversarial examples.

\cite{freiesleben2022intriguing} propose two definitional differences
between Adversarial Examples (AE) and Counterfactual Explanations (CE):
firstly, and more importantly according to the authors, the term AE
implies missclassification, which is not the case for CE
(\textbf{\emph{this might be a useful notion for use to distinguish
between adversarials and explanations during training}}); secondly, they
argue that closeness plays a more critical role in the context of CE but
confess that even counterfactuals that are not close might be relevant
explanations. \cite{pawelczyk2022exploring} show that CE and AE are
equivalent under certain conditions and derive upper bounds on the
distances between them.

\subsection{Closely Related}\label{closely-related}

\cite{guo2023counternet} are the first to propose end-to-end training
pipeline that includes counterfactual explanations as part of the
training prodeduce. In particular, they propose a specific network
architecture that includes a predictor and CE generator network
(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
network are learnable. Counterfactuals are generated during each
training iteration and fed back to the predictor network
(\textbf{\emph{here we are aligned}}). In contrast, we impose no
restrictions on the neural network architecture at all.
(\textbf{\emph{to ensure the one-hot encoding of categorical features is
maintained, they simple use softmax (might be interesting for CE.jl)}})
Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness. (\textbf{\emph{because partial gradients with
respect to the classification loss component and the counterfactual
validity component point in opposite directions}}). To mitigate these
issues, the authors use block-wise gradient descent: they first update
with respect to classification loss and then use a second update with
respect to the other loss components (\textbf{\emph{this might be useful
for our task as well}}). \cite{ross2021learning} propose a way to train
models that are guaranteed to provide recourse for individuals with high
probability. The approach builds on adversarial training
(\textbf{\emph{here we are aligned}}), where in this context adversarial
examples are actively encouraged to exist, but only target attacks with
respect to the positive class. The proposed method allows for imposing a
set of actionable recourse ex-ante: for example, users can impose
mutability constraints for features (\textbf{\emph{here we are
aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})

\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT) with intention of improving generalization and robustness
of language models. Specifically, they propose to proceed as follows:
firstly, identify training samples that are subject to high predictive
uncertainty (entropy); secondly, generate counterfactual explanations
for those samples; and, finally, finetune the model on the augmented
dataset that includes the generated counterfactuals.

@@

>>>>

\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed desirable with
respect to the underlying data and stakeholder requirements. Recent work
has shown that under this premise, the task of learning desirable
explanations is effectively reassigned from the model itself to the
(post-hoc) counterfactual explainer. Building on that work, we propose a
novel model objective that leverages counterfactuals during the training
phase (ad-hoc) in order to minimize the divergence between learned
representations and desirable explanations. Through extensive
experiments, we demonstrate that our proposed methodology facilitates
training models that inherently deliver desirable explanations while
maintaining high predictive performance.
\keywords{Counterfactual Explanations \and Explainable
AI \and Representation Learning}
\end{abstract}

@@

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, \cite{szegedy2013intriguing} showed that artificial neural
networks (ANN) are sensitive to adversarial examples: counterfactuals of
model inputs that yield vastly different model predictions despite being
``imperceptible'' in that they are semantically indifferent from their
factual counterparts. Despite partially effective mitigation strategies
such as \textbf{adversarial training} \cite{goodfellow2014explaining},
truly robust deep learning (DL) remains unattainable even for models
that are considered shallow by today's standards
\cite{kolter2023keynote}.

@@

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems (ADMS) involving
opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
in monitoring models and providing recourse to individuals to turn
adverse outcomes (e.g.~``loan application rejected'') into positive ones
(``application accepted''). \cite{wachter2017counterfactual} propose
\textbf{counterfactual explanations} as an effective approach to achieve
this: they explain how factual inputs need to change in order for some
fitted model to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed desirable by humans.
Instead, the choice has typically been to improve the capacity of XAI
tools to identify the subset explanations that are both desirable and
valid for any given model, independent of whether the learned
representations are also compatible with undesirable explanations
\cite{altmeyer2024faithful}. Fortunately, recent findings indicate that
explainability can arise as byproduct of regularization techniques aimed
at other objectives such as robustness, generalization and generative
capacity \cite{schut2021generating}.

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed desirable with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tieing the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some undesirable explanations from the solution
space.

\subsection{Adversarial Examples are Counterfactual
Explanations}\label{adversarial-examples-are-counterfactual-explanations}

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization. \cite{teney2020learning}
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on, firstly, identifying similar input samples with different
annotations and, secondly, ensuring that the gradient of the classifier
aligns with the vector between pairs of counterfactual inputs using the
cosine distance as a loss function. In the natural language processing
(NLP) domain, counterfactuals have similarly been used to improve models
through data augmentation: \cite{wu2021polyjuice}, propose
\emph{POLYJUICE}, a general-purpose counterfactual generator for
language models. They demonstrate empirically that augmenting training
data through \emph{POLYJUICE} counterfactuals improves robustness in a
number of NLP tasks. \cite{luu2023counterfactual} introduce
Counterfactual Adversarial Training (CAT), which also aims at improving
generalization and robustness of language models. Specifically, they
propose to proceed as follows: firstly, they identify training samples
that are subject to high predictive uncertainty; secondly, they generate
counterfactual explanations for those samples; and, finally, they
fine-tune the given language model on the augmented dataset that
includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations (CE) and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problem in representation learning. Formally,
AE can be seen as the subset of CE, for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
certain conditions and derive theoretical upper bounds on the distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals to move from an adverse
outcome to some positive target class with high probability. The
approach proposed by \cite{ross2021learning} builds on adversarial
training, where in this context susceptibility to targeted adversarial
examples for the positive class is explicitly induced. The proposed
method allows for imposing a set of actionability constraints ex-ante:
for example, users can specify that certain features (e.g.~\emph{age},
\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
are the first to propose an end-to-end training pipeline that includes
counterfactual explanations as part of the training procedure. In
particular, they propose a specific network architecture that includes a
predictor and CE generator network, where the parameters of the CE
generator network are learnable. Counterfactuals are generated during
each training iteration and fed back to the predictor network. In
contrast to \cite{guo2023counternet}, we impose no restrictions on the
neural network architecture at all.

\subsection{Beyond Robustness}\label{beyond-robustness}

Improving the adversarial robustness of models is not the only path
towards aligning representations with desirable explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally \textbf{faithful} to the model in that they are consistent
with what the model has learned about the underlying data. Formally,
they rely on tools from energy-based modelling to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields desirable (or
\textbf{plausible}) explanations if and only if the underlying model has
learned representations that align with them. They find that both deep
ensembles \cite{lakshminarayanan2016simple} and joint energy-based
models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating desirable
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with desirable explanations and has inspired our
proposed counterfactual training objective, as we explain in
Section~\ref{sec-method}.

@@

\begin{definition}[Model
Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}

~

\end{definition}
